SimBio: a flexible and extensible Python package for systems modelling

### 2023-11-10 article.qmd
Extend introduction

Chemical Reaction Networks (CRN) models
species and rates
Several modalities (ODE, SDE and jump process)
writing equations directly is error-prone and not easily reusable
Several packages exist to simulate them

Biosimulators.org, a registry of simulation tools, list at least 15 Python softwares.
(some) wrappers to external libraries
script the model definition
avoid exporting to csv and loading in Python
non-extendable from python
large community

Writing models
GUI (e.g., COPASI)
DSL (e.g, Antimony)
Python (e.g, PySB)
Benefits of IDE integration (syntax highlighting, etc)

Here we present SimBio, a Python package
that leverages standard modern Python syntax to integrate well with IDEs and static analysis tools.
It is written in pure Python such that it can be easily introspected and extended from Python.

It focused on providing a user experience, maintainability, and scalability

Clean API to create tools (graphs, tables, etc)

>>>>

Chemical Reaction Networks (CRN)
are the basis of many models used across the biological sciences and epidemiology.
They consist of a set of species and reactions that mediate conversions between them.
Several modalities exist to simulate these systems:
deterministic Ordinary Differential Equations (ODE) which model a macroscopic behavior,
Stochastic Differential Equations (SDE) which model microscopic fluctuations,
and jump processes (Gillespie-like simulations) which take into account the discreteness of the populations.
Hence, it is useful to write these models in a higher level description
which can be reused for these different types of simulations.
equations directly is error-prone and not easily reusable

Several tools already exist to define and simulate these systems.
In particular,
Biosimulators.org,
a registry of simulation tools,
list at least 15 Python softwares
among which are included
COPASI,
Tellurium,
GillesPy2
PySB.
However,
some are only Python wrappers to external (C, C++) libraries.
They allow to "control" the model definition and simulation in a Python script,
but it is not possible to use existing Python packages
to change the way the simulation is carried out.
In other words, they are not extensible from Python.

avoid exporting to csv and loading in Python
large community

To write these models,
there are three main approaches.
Tools such as COPASI provide a Graphical User Interface (GUI),
where the components, relations and their values are defined.
Others use a Domain Specific Language (DSL),
such as Antimony,
which is a less verbose version of the System Biology Markup Language (SBML).
The last approach is to use a programming language such as Python,
which is the approach used by PySB.

Benefits of IDE integration (syntax highlighting, etc)

Here we present SimBio, a Python package to define and simulate reaction networks.
From the end-user point of view,
it is focused on providing an ergonomic experience
by integrating well with IDEs and static analysis tools
through the use of standard modern Python syntax.
From the power-user point of view,
it can be extended from Python,
since its coded in pure Python.
Finally,
from the developer/maintainer point of view,
it has a modular architecture with a clear separation of concerns,
making it easier to maintain or to contribute new code.

the models can be easily introspected,
and simulation extended from Python.
Clean API to create tools (graphs, tables, etc)
